---
title: "stats506_hw1"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#questions for prof
#didn't reassign variable names until table part because i didn't notice, should i go back and change it?
```

```{r}
#install.packages("gt")
#install.packages("rstatix")
#install.packages("roxygen2")
#install.packages("schoolmath")
```

```{r}
library(tidyverse)
library(gt)
library(rstatix) #t_test
library(roxygen2)
library(schoolmath) #is.positive
```

Problem 1 - Exploratory Analysis of Soybean Cultivars

a.  import data

```{r}
#load data
soybean <- read.csv("data.csv")
head(soybean)
```

b.  verify information

```{r}
#number of observations in soybean dataset
num_obs_soybean <- nrow(soybean)
num_obs_soybean

#number of unique 'vals' in Cultivar variable 
#unique: duplicates removed, length: count of elements in unique(), 
num_Cultivar <- length(unique(soybean$Cultivar))
num_Cultivar

#check seasons
unique(soybean$Season)

#table showing num of obs in each season
num_obs_byseason <- soybean |> 
  group_by(Season) |>
  summarize("Number of observations" = n())
num_obs_byseason
```

c.  use data to answer questions

<!-- -->

1.  What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
#correlation between ph and gy
ph_gy_corr <- cor(soybean$PH, soybean$GY)
ph_gy_corr
```

The correlation between plant height and grain yield is about 0.123

2.  Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
#correlation between ph and gy by season
#sort by season first then do cor()
corr_byseason <- soybean |>
  group_by(Season) |>
  summarize(Correlation = cor(PH, GY))
corr_byseason
```

Season 1 has the higher absolute correlation of 0.23 compared to season 2's 0.023.

3.  What is the grain yield of the soybean observation with the greatest plant height?

```{r}
#get index of tallest, get GY using index -> GY of tallest plant
#https://r-statistics.co/base-max-in-R.html used to find which.max() func
tallest_idx <- which.max(soybean$PH) #index  of largest val
tallest_idx
tallest_gy <- soybean$GY[tallest_idx]
tallest_gy
```

The grain yield of the tallest soybean plant is 3304.678 kg/ha.

4.  What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
#get MHG avg
avg_MHG <- mean(soybean$MHG)

#df with number of obs above avg and % obs above avg (for MHG)
#filter for above avg MHG then calc %
above_avg_MHG <- soybean |>
  filter(MHG > avg_MHG) |>
  summarize(
    #number of obs above avg
    above_avg_count = n(),
    #percentage of obs above avg
    #original count from part b (num_obs_soybean)
    above_avg_percent = (above_avg_count/num_obs_soybean)*100)
above_avg_MHG
```

About 47.19% of observations have a thousand seed weight greater than the overall average.

5.  Which cultivar has the highest average grain yield?

```{r}
#get df with average GY for each cultivar 
avg_gy_bycultivar <- soybean |> 
  group_by(Cultivar) |>
  summarize(avg_gy = mean(GY))
avg_gy_bycultivar

#use which.max to get index of highest avg GY then use that to get cultivar
avg_gy_bycultivar$Cultivar[which.max(avg_gy_bycultivar$avg_gy)]
```

The cultivar with the highest average grain yield is 98R30 CE.

d.  create table showing averages

<!-- -->

1.  One row containing overall averages.

```{r}
#average all continuous variables (PH, etc)
#https://dplyr.tidyverse.org/reference/summarise_all.html, https://dplyr.tidyverse.org/reference/across.html; examples for doing multiple operations at once
soybean_overall_avgs <- soybean |>
  #.x: current column, didn't include na.rm (these cols don't have missing vals)
  summarize(across(-c(Season, Cultivar, Repetition), ~mean(.x)))
soybean_overall_avgs
```

2.  One row for each season containing season-specific averages.

```{r}
#avg of cont vars by season
#first group by season
soybean_season_avgs <- soybean |>
  group_by(Season) |>
  #then get averages of each column (by season)
  summarize(across(-c(Cultivar, Repetition), ~mean(.x)))
soybean_season_avgs
```

Combined table:

```{r}
#create season column (put in front) for overall avgs to help with merging dfs
soybean_overall_avgs <- soybean_overall_avgs |>
  mutate(Season = "All", .before = "PH")

#new df with overall avgs and avgs by season
soybean_avgs_df <- rbind(soybean_overall_avgs, soybean_season_avgs)
#change season variable values to strings for clarity
soybean_avgs_df <- soybean_avgs_df |>
  mutate(
    Season = case_when(
      Season == "All" ~ "Averages for all seasons", 
      Season == 1 ~ "Averages for season 1",
      Season == 2 ~ "Averages for season 2",
      TRUE ~ NA_character_
    )
  )

#table; got idea for making table from: https://gt.rstudio.com/, https://www.rdocumentation.org/packages/gt/versions/0.8.0/topics/gt
soybean_avgs_table <- soybean_avgs_df |>
  gt() |>
  tab_header(title = "Overall and By-Season Averages") |> 
  #learned functions for cleaning up table from: https://r-statistics.co/gt-Package.html
  cols_label(
    PH = "Plant Height",
    IFP = "Insertion of first pod", 
    NLP = "Number of stems",
    NGP = "Number of legumes per plant",
    NGL = "Number of grains per plant",
    NS = "Number of grains per pod",
    MHG = "Thousand seed weight",
    GY = "Grain yield"
    ) |> 
  cols_align(align = "center")
soybean_avgs_table
```


e.  Carry out a series of t-tests to examine whether the grain yield (GY) differs between the two seasons. State the null and the alternative hypotheses, present R output and interpret results.


Null hypothesis: grain yield (GY) doesn't differ between season 1 and season 2 for cultivar X

Alternative hypothesis: grain yield (GY) differs between season 1 and season 2 for cultivar X

X = whichever cultivar is being filtered for (noted in comments)

```{r}
#soybean; note: data looks paired; 

#74K75RSF CE
cultivar1_test <- soybean |>
  filter(Cultivar == "74K75RSF CE") |> 
  t_test(GY ~ Season, paired = TRUE)
cultivar1_test

#77HO111I2X - GUAPORÉ	
cultivar2_test <- soybean |>
  filter(Cultivar == "77HO111I2X - GUAPORÉ") |> 
  t_test(GY ~ Season, paired = TRUE)
cultivar2_test

#79I81RSF IPRO	
cultivar3_test <- soybean |>
  filter(Cultivar == "79I81RSF IPRO") |> 
  t_test(GY ~ Season, paired = TRUE)
cultivar3_test

#82HO111 IPRO - HO COXIM IPRO	
cultivar4_test <- soybean |>
  filter(Cultivar == "82HO111 IPRO - HO COXIM IPRO") |> 
  t_test(GY ~ Season, paired = TRUE)
cultivar4_test

#82I78RSF IPRO
cultivar5_test <- soybean |>
  filter(Cultivar == "82I78RSF IPRO") |> 
  t_test(GY ~ Season, paired = TRUE)
cultivar5_test
```

Problem 2 - Adult Income Dataset

a, b. import data, clean variable names

```{r}
#import data
adult <- read_delim("adult.data", delim = ",", col_names = FALSE, show_col_types = FALSE, na = c(" ?", "NA"))
head(adult)

#change col names 
adult <- adult |>
  rename(
    Age = X1,
    Employment_stat = X2,
    fnlwgt = X3, #no explanation of this var's abbrv.
    Education = X4,
    Education_yrs = X5,
    Marital_stat = X6, 
    Occupation = X7,
    Relationship = X8, 
    Race = X9,
    Sex = X10,
    Capital_gain = X11,
    Capital_loss = X12, 
    Hrs_per_wk = X13,
    Native_country = X14,
    Income = X15
  )
head(adult)

#anyNA(adult) showed there were initially no NA vals, used method below:
#check what NA values look like using unique() (only checked cols that were said to have missing values in data description), then added variants into read_delim above
#unique(adult$Occupation)
#unique(adult$Employment_stat)
#unique(adult$Native_country)
```

c. examine missing vals, report num of missing vals for each var
```{r}
#across(): apply to multiple cols (can use in summarize), want to find na vals (is.na()) for each col/var, then summarize into table 
adult_na <- adult |>
  summarize(across(everything(), ~sum(is.na(.x))))
adult_na
```

Employment_stat (or workclass in the original dataset) has 1836 missing values, occupation has 1843 missing values, and native_country has 583 missing values. 

d. Restrict the sample to individuals working at least 10 hours per week. Confirm the sample size before and after filtering.
```{r}
#check sample size before filtering
num_obs_adult <- nrow(adult)
num_obs_adult

#checking variable type since initial filter gave 0 obs
class(adult$Hrs_per_wk) #character class -> need to change to numeric for filter() to work

#filter for people working >= 10 hrs/wk
adult_over_10 <- adult |>
  mutate(
    Hrs_per_wk = as.numeric(Hrs_per_wk),
  ) |>
  filter(Hrs_per_wk >= 10)
  
#check sample size after filtering 
num_obs_over_10 <- nrow(adult_over_10)
num_obs_over_10

```

Of the 32,561 observations in the original dataset, 32,103 individuals were working at least 10 hours a week. 

e + f + g. Remove observations with missing values for occupation, workclass, or native country. Explain why. Eliminate any observations with: age less than 18, age greater than 90, education level identified as Preschool. Report your final sample size.
```{r}
#assuming we're working with adjusted dataset from part d
adult_cleaned <- adult_over_10 |> 
  #remove obs with missing vals
  drop_na(Occupation, Employment_stat, Native_country) |>
  #keep everything between 18 and 90
  filter(Age >= 18 & Age <= 90) |>
  #remove obs with education level = preschool 
  filter(Education != "Preschool")
```

The description of the original dataset specified that occupation, workclass, and native country were the only three variables with missing values, and removing them helps prevent issues later down the line if we want to do data analysis. 

g. Report your final sample size
```{r}
#check final sample size 
final_sample_size <- nrow(adult_cleaned)
final_sample_size
```

The final sample size is 29,568. 

Problem 3 - Happy Numbers

a. Write a function nextHappy() that given a positive integer, computes the next value in the happy-number sequence. Be sure to provide a reasonable error on an invalid input. Be sure to document your function.
Input: A positive integer
Output: A positive integer
Demonstrate it works by reproducing the examples.
```{r}
#' Function to compute the next value in the happy-number sequence give a positive integer
#' 
#' @param pos_int a positive integer 
#' @return the next happy-number seq value 
nextHappy <- function(pos_int) {
  #got idea to use %%1 from here for sanity checks: https://stackoverflow.com/questions/3476782/check-if-the-number-is-integer
  #sanitize pos_int
  #if it's not positive or the decimal part doesn't = 0, then stop func + display error message
  if (!is.positive(pos_int) | (pos_int %% 1) != 0) {
    stop("pos_int must be a positive integer ")
  }
  
  #empty vector to store squared values in
  squares <- c()
  
  #loop through each digit of the pos_int
  while(pos_int != 0) {
    #get unit digit
    digit <- pos_int %% 10
    #square unit digit
    digit_squared <- digit^2
    #add squared value to vector
    squares <- append(squares, digit_squared)
    #update pos_int to get rid of unit digit
    pos_int <- floor(pos_int / 10)
  }
  
  #get sum of all the squared digits
  sum_of_squares <- sum(squares)
  #print final sum a.k.a. next happy-number sequence value
  sum_of_squares
}

nextHappy(19)
nextHappy(82)
```

b. Create a function happySequence() that returns the sequence beginning at a given number. 

Starting with n, replace it with the sum of the squares of its digits, and repeat the process until n equals 1, or it loops endlessly in a cycle that does not include 1. Those numbers for which this process ends in 1 are Happy Numbers, while those that do not end in 1 are unhappy numbers.

Use your nextHappy() function to perform the calculation. Be sure to provide a reasonable error on an invalid input. Be sure to document your function (see instructions above).

Input: A positive integer

Output: A list containing: 

1. sequence: A vector containing the values in the Happy sequence, beginning with the input value and ending at 1 if the number is happy.

2. isHappy: A logical value indicating whether the input is a happy number (TRUE if the sequence reaches 1, FALSE otherwise). Demonstrate it works by reproducing the examples.
```{r}
#' Function to see if a given value is a happy number or unhappy number and return a Happy sequence if it is a happy number
#' 
#' @param pos_int a positive integer 
#' @return a logical value that indicates if the input n is a happy number or unhappy number, and a vector with the values in the Happy sequence if n is a happy number beginning with n and ending with 1
happySequence <- function(n) {
  #sanitize n
  #if it's not positive or the decimal part doesn't = 0, then stop func + display error message
  if (!is.positive(n) | (n %% 1) != 0) {
    stop("pos_int must be a positive integer ")
  }
  
  #vector for sequence starting with initial input value
  Happy_sequence = c(n)
  
  #if the initial input is 1, return the sequence 
  if (n == 1) {
    return(list(sequence = Happy_sequence, isHappy = TRUE))
  } else {
    #got idea for loop from happy numbers wiki page (https://en.wikipedia.org/wiki/Happy_number): check for repeating value
    #as long as n doesn't reach 1 or repeat a value, continue with sequence 
    while (n != 1) {
      n <- nextHappy(n)
      #add new value to sequence vector
      Happy_sequence <- append(Happy_sequence, n)
      
      #if the latest value is equal to 1, then it's a happy number, return that and the sequence
      if (Happy_sequence[length(Happy_sequence)] == 1) {
        return_list = list(sequence = Happy_sequence, isHappy = TRUE)
        return(return_list)
      #if there's a duplicate in the sequence (minus the latest appended value), then it's an unhappy number
      } else if (n %in% Happy_sequence[-length(Happy_sequence)]) {
        return(isHappy = FALSE)
      }
    }
  }
}

#examples
happySequence(19)
happySequence(4)
happySequence(1)
```

c. Among integers 1 through 1000: How many are happy? What percentage are happy? What starting value produces the longest happy sequence before terminating or cycling?
```{r}

```




