---
title: "stats506_hw1"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#questions for prof
#didn't reassign variable names until table part because i didn't notice, should i go back and change it?
#t test part in problem 1

```

```{r}
#install.packages("gt")
#install.packages("rstatix")
```

```{r}
library(tidyverse)
library(gt)
library(rstatix)
```

Problem 1 - Exploratory Analysis of Soybean Cultivars

a.  import data

```{r}
#load data
soybean <- read.csv("data.csv")
head(soybean)
```

b.  verify information

```{r}
#number of observations in soybean dataset
num_obs_soybean <- nrow(soybean)
num_obs_soybean

#number of unique 'vals' in Cultivar variable 
#unique: duplicates removed, length: count of elements in unique(), 
num_Cultivar <- length(unique(soybean$Cultivar))
num_Cultivar

#check seasons
unique(soybean$Season)

#table showing num of obs in each season
num_obs_byseason <- soybean |> 
  group_by(Season) |>
  summarize("Number of observations" = n())
num_obs_byseason
```

c.  use data to answer questions

<!-- -->

1.  What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
#correlation between ph and gy
ph_gy_corr <- cor(soybean$PH, soybean$GY)
ph_gy_corr
```

The correlation between plant height and grain yield is about 0.123

2.  Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
#correlation between ph and gy by season
#sort by season first then do cor()
corr_byseason <- soybean |>
  group_by(Season) |>
  summarize(Correlation = cor(PH, GY))
corr_byseason
```

Season 1 has the higher absolute correlation of 0.23 compared to season 2's 0.023.

3.  What is the grain yield of the soybean observation with the greatest plant height?

```{r}
#get index of tallest, get GY using index -> GY of tallest plant
#https://r-statistics.co/base-max-in-R.html used to find which.max() func
tallest_idx <- which.max(soybean$PH) #index  of largest val
tallest_idx
tallest_gy <- soybean$GY[tallest_idx]
tallest_gy
```

The grain yield of the tallest soybean plant is 3304.678 kg/ha.

4.  What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
#get MHG avg
avg_MHG <- mean(soybean$MHG)

#df with number of obs above avg and % obs above avg (for MHG)
#filter for above avg MHG then calc %
above_avg_MHG <- soybean |>
  filter(MHG > avg_MHG) |>
  summarize(
    #number of obs above avg
    above_avg_count = n(),
    #percentage of obs above avg
    #original count from part b (num_obs_soybean)
    above_avg_percent = (above_avg_count/num_obs_soybean)*100)
above_avg_MHG
```

About 47.19% of observations have a thousand seed weight greater than the overall average.

5.  Which cultivar has the highest average grain yield?

```{r}
#get df with average GY for each cultivar 
avg_gy_bycultivar <- soybean |> 
  group_by(Cultivar) |>
  summarize(avg_gy = mean(GY))
avg_gy_bycultivar

#use which.max to get index of highest avg GY then use that to get cultivar
avg_gy_bycultivar$Cultivar[which.max(avg_gy_bycultivar$avg_gy)]
```

The cultivar with the highest average grain yield is 98R30 CE.

d.  create table showing averages

<!-- -->

1.  One row containing overall averages.

```{r}
#average all continuous variables (PH, etc)
#https://dplyr.tidyverse.org/reference/summarise_all.html, https://dplyr.tidyverse.org/reference/across.html; examples for doing multiple operations at once
soybean_overall_avgs <- soybean |>
  #.x: current column, didn't include na.rm (these cols don't have missing vals)
  summarize(across(-c(Season, Cultivar, Repetition), ~mean(.x)))
soybean_overall_avgs
```

2.  One row for each season containing season-specific averages.

```{r}
#avg of cont vars by season
#first group by season
soybean_season_avgs <- soybean |>
  group_by(Season) |>
  #then get averages of each column (by season)
  summarize(across(-c(Cultivar, Repetition), ~mean(.x)))
soybean_season_avgs
```

Combined table:

```{r}
#create season column (put in front) for overall avgs to help with merging dfs
soybean_overall_avgs <- soybean_overall_avgs |>
  mutate(Season = "All", .before = "PH")

#new df with overall avgs and avgs by season
soybean_avgs_df <- rbind(soybean_overall_avgs, soybean_season_avgs)
#change season variable values to strings for clarity
soybean_avgs_df <- soybean_avgs_df |>
  mutate(
    Season = case_when(
      Season == "All" ~ "Averages for all seasons", 
      Season == 1 ~ "Averages for season 1",
      Season == 2 ~ "Averages for season 2",
      TRUE ~ NA_character_
    )
  )

#table; got idea for making table from: https://gt.rstudio.com/, https://www.rdocumentation.org/packages/gt/versions/0.8.0/topics/gt
soybean_avgs_table <- soybean_avgs_df |>
  gt() |>
  tab_header(title = "Overall and By-Season Averages") |> 
  #learned functions for cleaning up table from: https://r-statistics.co/gt-Package.html
  cols_label(
    PH = "Plant Height",
    IFP = "Insertion of first pod", 
    NLP = "Number of stems",
    NGP = "Number of legumes per plant",
    NGL = "Number of grains per plant",
    NS = "Number of grains per pod",
    MHG = "Thousand seed weight",
    GY = "Grain yield"
    ) |> 
  cols_align(align = "center")
soybean_avgs_table
```

e.  Carry out a series of t-tests to examine whether the grain yield (GY) differs between the two seasons. State the null and the alternative hypotheses, present R output and interpret results.

Overall:

Null hypothesis: grain yield (GY) doesn't differ between season 1 and season 2

Alternative hypothesis: grain yield (GY) differs between season 1 and season 2

```{r}
#soybean; data looks paired
soybean |> 
  t_test(GY ~ Season, paired = TRUE)
```

Problem 2 - Adult Income Dataset

a, b. import data, clean variable names

```{r}
#import data
adult <- read_delim("adult.data", delim = ",", col_names = FALSE, show_col_types = FALSE, na = c(" ?", "NA"))
head(adult)

#change col names 
adult <- adult |>
  rename(
    Age = X1,
    Employment_stat = X2,
    fnlwgt = X3, #no explanation of this var's abbrv.
    Education = X4,
    Education_yrs = X5,
    Marital_stat = X6, 
    Occupation = X7,
    Relationship = X8, 
    Race = X9,
    Sex = X10,
    Capital_gain = X11,
    Capital_loss = X12, 
    Hrs_per_wk = X13,
    Native_country = X14,
    Income = X15
  )
head(adult)

#anyNA(adult) showed there were initially no NA vals, used method below:
#check what NA values look like using unique() (only checked cols that were said to have missing values in data description), then added variants into read_delim above
#unique(adult$Occupation)
#unique(adult$Employment_stat)
#unique(adult$Native_country)
```

c. examine missing vals, report num of missing vals for each var
```{r}
#across(): apply to multiple cols (can use in summarize), want to find na vals (is.na()) for each col/var, then summarize into table 
adult_na <- adult |>
  summarize(across(everything(), ~sum(is.na(.x))))
adult_na
```

Employment_stat (or workclass in the original dataset) has 1836 missing values, occupation has 1843 missing values, and native_country has 583 missing values. 

d. Restrict the sample to individuals working at least 10 hours per week. Confirm the sample size before and after filtering.
```{r}
#check sample size before filtering
num_obs_adult <- nrow(adult)
num_obs_adult

#checking variable type since initial filter gave 0 obs
class(adult$Hrs_per_wk) #character class -> need to change to numeric for filter() to work

#filter for people working >= 10 hrs/wk
adult_over_10 <- adult |>
  mutate(
    Hrs_per_wk = as.numeric(Hrs_per_wk),
  ) |>
  filter(Hrs_per_wk >= 10)
  
#check sample size after filtering 
num_obs_over_10 <- nrow(adult_over_10)
num_obs_over_10

```

Of the 32,561 observations in the original dataset, 32,103 individuals were working at least 10 hours a week. 

e + f + g. Remove observations with missing values for occupation, workclass, or native country. Explain why. Eliminate any observations with: age less than 18, age greater than 90, education level identified as Preschool. Report your final sample size.
```{r}
#assuming we're working with adjusted dataset from part d
adult_cleaned <- adult_over_10 |> 
  #remove obs with missing vals
  drop_na(Occupation, Employment_stat, Native_country) |>
  #keep everything between 18 and 90
  filter(Age >= 18 & Age <= 90) |>
  #remove obs with education level = preschool 
  filter(Education != "Preschool")
```

The description of the original dataset specified that occupation, workclass, and native country were the only three variables with missing values, and removing them helps prevent issues later down the line if we want to do data analysis. 

g. Report your final sample size
```{r}
#check final sample size 
final_sample_size <- nrow(adult_cleaned)
final_sample_size
```

The final sample size is 29,568. 









