---
title: "stats506_hw1"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#questions for prof
#didn't reassign variable names until table part because i didn't notice, should i go back and change it?
#t test part in problem 1

```

```{r}
#install.packages("gt")
#install.packages("rstatix")
```

```{r}
library(tidyverse)
library(gt)
library(rstatix)
```

Problem 1 - Exploratory Analysis of Soybean Cultivars

a.  import data

```{r}
#load data
soybean <- read.csv("data.csv")
head(soybean)
```

b.  verify information

```{r}
#number of observations in soybean dataset
num_obs_soybean <- nrow(soybean)
num_obs_soybean

#number of unique 'vals' in Cultivar variable 
#unique: duplicates removed, length: count of elements in unique(), 
num_Cultivar <- length(unique(soybean$Cultivar))
num_Cultivar

#check seasons
unique(soybean$Season)

#table showing num of obs in each season
num_obs_byseason <- soybean |> 
  group_by(Season) |>
  summarize("Number of observations" = n())
num_obs_byseason
```

c.  use data to answer questions

<!-- -->

1.  What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
#correlation between ph and gy
ph_gy_corr <- cor(soybean$PH, soybean$GY)
ph_gy_corr
```

The correlation between plant height and grain yield is about 0.123

2.  Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
#correlation between ph and gy by season
#sort by season first then do cor()
corr_byseason <- soybean |>
  group_by(Season) |>
  summarize(Correlation = cor(PH, GY))
corr_byseason
```

Season 1 has the higher absolute correlation of 0.23 compared to season 2's 0.023.

3.  What is the grain yield of the soybean observation with the greatest plant height?

```{r}
#get index of tallest, get GY using index -> GY of tallest plant
#https://r-statistics.co/base-max-in-R.html used to find which.max() func
tallest_idx <- which.max(soybean$PH) #index  of largest val
tallest_idx
tallest_gy <- soybean$GY[tallest_idx]
tallest_gy
```

The grain yield of the tallest soybean plant is 3304.678 kg/ha.

4.  What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
#get MHG avg
avg_MHG <- mean(soybean$MHG)

#df with number of obs above avg and % obs above avg (for MHG)
#filter for above avg MHG then calc %
above_avg_MHG <- soybean |>
  filter(MHG > avg_MHG) |>
  summarize(
    #number of obs above avg
    above_avg_count = n(),
    #percentage of obs above avg
    #original count from part b (num_obs_soybean)
    above_avg_percent = (above_avg_count/num_obs_soybean)*100)
above_avg_MHG
```

About 47.19% of observations have a thousand seed weight greater than the overall average.

5.  Which cultivar has the highest average grain yield?

```{r}
#get df with average GY for each cultivar 
avg_gy_bycultivar <- soybean |> 
  group_by(Cultivar) |>
  summarize(avg_gy = mean(GY))
avg_gy_bycultivar

#use which.max to get index of highest avg GY then use that to get cultivar
avg_gy_bycultivar$Cultivar[which.max(avg_gy_bycultivar$avg_gy)]
```

The cultivar with the highest average grain yield is 98R30 CE.

d.  create table showing averages

<!-- -->

1.  One row containing overall averages.

```{r}
#average all continuous variables (PH, etc)
#https://dplyr.tidyverse.org/reference/summarise_all.html; examples for doing multiple operations at once
soybean_overall_avgs <- soybean |>
  #.x: current column, didn't include na.rm (these cols don't have missing vals)
  summarize(across(-c(Season, Cultivar, Repetition), ~mean(.x)))
soybean_overall_avgs
```

2.  One row for each season containing season-specific averages.

```{r}
#avg of cont vars by season
#first group by season
soybean_season_avgs <- soybean |>
  group_by(Season) |>
  #then get averages of each column (by season)
  summarize(across(-c(Cultivar, Repetition), ~mean(.x)))
soybean_season_avgs
```

Combined table:

```{r}
#create season column (put in front) for overall avgs to help with merging dfs
soybean_overall_avgs <- soybean_overall_avgs |>
  mutate(Season = "All", .before = "PH")

#new df with overall avgs and avgs by season
soybean_avgs_df <- rbind(soybean_overall_avgs, soybean_season_avgs)
#change season variable values to strings for clarity
soybean_avgs_df <- soybean_avgs_df |>
  mutate(
    Season = case_when(
      Season == "All" ~ "Averages for all seasons", 
      Season == 1 ~ "Averages for season 1",
      Season == 2 ~ "Averages for season 2",
      TRUE ~ NA_character_
    )
  )

#table; got idea for making table from: https://gt.rstudio.com/, https://www.rdocumentation.org/packages/gt/versions/0.8.0/topics/gt
soybean_avgs_table <- soybean_avgs_df |>
  gt() |>
  tab_header(title = "Overall and By-Season Averages") |> 
  #learned functions for cleaning up table from: https://r-statistics.co/gt-Package.html
  cols_label(
    PH = "Plant Height",
    IFP = "Insertion of first pod", 
    NLP = "Number of stems",
    NGP = "Number of legumes per plant",
    NGL = "Number of grains per plant",
    NS = "Number of grains per pod",
    MHG = "Thousand seed weight",
    GY = "Grain yield"
    ) |> 
  cols_align(align = "center")
soybean_avgs_table
```

e.  Carry out a series of t-tests to examine whether the grain yield (GY) differs between the two seasons. State the null and the alternative hypotheses, present R output and interpret results.

Overall:

Null hypothesis: grain yield (GY) doesn't differ between season 1 and season 2

Alternative hypothesis: grain yield (GY) differs between season 1 and season 2

```{r}
#soybean; data looks paired
soybean |> 
  t_test(GY ~ Season, paired = TRUE)
```

Problem 2 - Adult Income Dataset

a.  

```{r}
#import data
adult <- read_delim("adult.data")
head(adult)

#clean up 
#adult <- adult |>
  
```
